Data Specialization for a Staged Lambda Calculus

### 2015-07-06 paper/splitting.tex
added proof cases

Partial value splitting (given by the relation $\vsplito$) converts a 
partial value $v$ into a split value $\mval i q$.
The first component of the split value, the \emph{immediate value} $i$, is a monostage value holding
the result of all first-stage components of $v$.  The second
component, $q$, is a residual representing all second-stage
variable references of $v$.

@@

\TODO do some cases of the proof, and use the opportunity to discuss why splitting does what it does

\paragraph{\bbtwo\ Compatility.}
\paragraph{\bbone\ Compatility.}
\paragraph{Functions.}
\paragraph{Next and prev.}
\paragraph{Ground and Caseg.}

@@

\input{figures/splitting-misc}
\input{figures/splitting-term}

>>>>

Partial value splitting (given by the relation $\vsplito$) converts a 
partial value $v$ into a split value $\mval i q$.
The first component, the \emph{immediate value} $i$, is a monostage value holding
the result of all first-stage components of $v$.  The second
component, $q$, is a residual representing all second-stage
variable references of $v$.

@@

In this section, we prove the \ref{lem:oneCorrect} and \ref{lem:twoCorrect} lemmas
by considering a few exemplar cases.

\paragraph{Reduction of {\tt pi1}.}
Define $E = \pio {\exv{\tup {v_1,v_2}}}$ and  
$C = \letin{\tup{y,z}}{\exv{\tup{\tup {i_1,i_2},\tup{}}}}{\tup{\pio {\exv y},\exv z}}$.
We are given $\stepmix  E {\exv {v_1}}$,
$\splitone E A C {l.\pio {\exv {\tup{q_1,q_2}}}}$, and
$\splitone {\exv {v_1}} A {\exv {\tup{i_1,\tup{}}}} {\_.\exv{q_1}}$,
and we need to show $\reduce C {\tup {i_1,\tup{}}}$
and $\pio {\exv {\tup{q_1,q_2}}} \equiv q_1$.
Both of these can be derived directly.
This pattern, where the outputs can be directly derived, extends to the {\tt pi2} and {\tt prev} reduction rules and all value marking rules.

\paragraph{Reduction of Application.} Let $E = \app {\exv{\fix f x e}} {\exv v}$,
$E' = [\fix fxe, v/f,x] e$, \\ 
$I = \fix f x {\letin {\tup{x,y}} c {\tup{x,\roll y}}}$, \\
$Q = \fix f {\tup{x,\roll l}} r$, and \\
$C = \left[
\talllet{\tup{y_1,z_1}}{\exv {\tup {I,\tup{}}}}{
\talllet{\tup{y_2,z_2}} {\exv {\tup {i_1,\tup{}}}}{
\talllet{\tup{y_3,z_3}}{\app{\exv {y_1}}{\exv {y_2}}}{\exv{\tup{y_3,\tup{z_1,z_2,z_3}}}\ttrpar\ttrpar\ttrpar}}}\right]$.
\\
We are given $\stepmix E {E'}$,
$\splitone E A C {\tup{\_,\_,l}.\app{\exv Q}{\exv{\tup{q_1,l}}}}$,
$\splitone {E'} A {[I,i_1/f,x]c} {l.[Q,q_1/f,x]r}$,
and $\reduce {[I,i_1/f,x]c} {\tup{i,b}}$.
From this, we can directly derive $\reduce C {\tup{i,\tup {\tup{},\tup{},\roll b}}}$
and $\app{\exv Q}{\exv{\tup{q_1,\roll b}}} \equiv [Q,q_1,b/f,x,l] r$.
This pattern applies to all of the other reduction rules involving substitution,
namely those for {\tt caseg}, and {\tt letg}. 
Observe that, in our formal version of the language, 
{\tt caseg} is the only construct where information is added to the boundary value.

\paragraph{Compatility of {\tt pi1} at \bbtwo.}
We are given 
$\steptwo {\pio e} {\pio {e'}}$, 
$\splittwo {\pio e} A p l {\pio r}$, 
$\splittwo {\pio {e'}} A {p'} {l'} {\pio {r'}}$,
and $\reduce {p'} {b'}$.
Inversion of the first three yields
$\steptwo e {e'}$, $\splittwo e A p l r$, and $\splittwo {e'} A {p'} {l'} {r'}$.
Using \ref{lem:twoCorrect} inductively gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r$.
From this, $[b/l]\pio r \equiv [b'/l']\pio {r}$ can be derived directly.
This pattern generalizes to all world \bbtwo\ compatibility rules.

\paragraph{Compatility of {\tt pi1} at \bbonem.}
We are given
$\stepmix {\pio e} {\pio {e'}}$, 
$\splitone {\pio e} A {\letin{\tup{y,z}}{c}{\tup{\pio {\exv y},\exv z}}} {l.\pio r}$, \\
\mbox{$\splitone {\pio {e'}} A {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {l'.\pio {r'}}$},
and \\ $\reduce {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {\tup{i,b'}}$.
Inversion of the first three yields 
$\stepmix e {e'}$, $\splitone e A c {l.r}$, and $\splitone {e'} A {c'} {l'.r'}$, 
and inversion of the reduction yields $\reduce {c'} {\tup{\tup{i,i_2},b'}}$.
Using \ref{lem:oneCorrect} inductively gives $\reduce c {\tup{\tup{i,i_2},b}}$ and $[b/l]r \equiv [b'/l']r$.
From this, we can derive $\reduce {\letin{\tup{y,z}}{c}{\tup{\pio {\exv y},\exv z}}} {\tup{i,b}}$ and 
$[b/l]\pio r \equiv [b'/l']\pio {r}$ directly.
This pattern generalizes to all world \bbonem\ compatibility rules.

\paragraph{Compatibility of {\tt gr}.}
We are given 
$\stepmix {\pure e} {\pure {e'}}$, \\
$\splitone {\pure e} A {\tup {e,\exv {\tup{}}}} {\_.\exv{\tup{}}}$, 
$\splitone {\pure {e'}} A {\tup {e',\exv {\tup{}}}} {\_.\exv{\tup{}}}$,
and $\reduce {\tup {e',\exv {\tup{}}}} {\tup{i,\tup{}}}$.
Inversion of the step and reduction yield \mbox{$\steppure e {e'}$} and $\reduce {e'} i$.
As a simple property of monostage reduction (which $\stepsym \bbonep$ is), we know $\reduce e i$.
From here, we can derive $\reduce {\tup {e,\exv {\tup{}}}} {\tup{i,\tup{}}}$ and $\exv{\tup {}} \equiv \exv{\tup {}}$ directly.

@@

\input{figures/splitting-term}
